---
title: "Analysis of Neuronal Data on Mice Decision Making"
author: "Laura Wang"
date: "March 8, 2023"
output:
 bookdown::html_document2:
    df_print: paged
    number_sections: False
    fig_caption: True
    code_folding: hide
    toc: TRUE
    toc_float:
      collapsed: true
      smooth_scroll: false
header-includes:
  \usepackage{caption}
---
```{r global_options, include=FALSE}
setwd(dirname(rstudioapi::getSourceEditorContext()$path)) #setting working directory
knitr::opts_chunk$set(fig.pos = 'H')
library(dplyr)
library(car)
library(MASS)
library(ggplot2)
library(knitr) #tables
library(moments) #skewness
library(kableExtra) #scroll box
library(cowplot)
library(lme4)
library(lmerTest)
library(gplots)
library(data.table) #melt
library(plotly)
library(lemon) #shift legend
library(gtable)
library(pROC)
library(caret)
library(lmtest) #likelihood ratio test
```
<style>
p.caption {
  font-size: 1.1em;
  font-weight: bold;
}

caption {
  color: black;
  font-weight: bold;
  font-size: 1.1em;
} 
</style>

```{r include=FALSE}
#function for moving legend position I found online. 

shift_legend2 <- function(p) {
  # ...
  # to grob
  gp <- ggplotGrob(p)
  facet.panels <- grep("^panel", gp[["layout"]][["name"]])
  empty.facet.panels <- sapply(facet.panels, function(i) "zeroGrob" %in% class(gp[["grobs"]][[i]]))
  empty.facet.panels <- facet.panels[empty.facet.panels]

  # establish name of empty panels
  empty.facet.panels <- gp[["layout"]][empty.facet.panels, ]
  names <- empty.facet.panels$name
  # example of names:
  #[1] "panel-3-2" "panel-3-3"

# now we just need a simple call to reposition the legend
  reposition_legend(p, 'center', panel=names)
}
```


# Abstract  

In this analysis, data from a neurological study done on mice was analyzed to determine whether or not the stimuli presented on the left and right side of a mouse interacted to create certain levels of neuronal activity in the visual cortex. The analysis showed that the left and right contrasts (stimuli) do in fact have an interaction effect and different combinations of these stimuli may produce different levels of activity. Additionally, a predictive model was built to try and use the measurements of neuron activity to predict the outcomes of some given trials. 

# Introduction  

This project focuses on understanding how neural activities in the visual cortex are related to the intake and processing of stimuli. The primary questions of interest are as follows:  

1. How do neurons in the visual cortex respond to the stimuli presented on the left and right?   

2. Can we predict the success or failure of each trial based on the activity of the neurons and the kinds of stimulus presented?  

In the first questions, the focus is on how the stimuli affect the neural activities. In particular, whether or not the stimuli have additive effects on the neural responses. To answer this, a 2-way ANOVA model with mixed effects will be fitted to the entire data set and the variances of a model with only additive effects and a model including an interaction term will be compared. To answer the second question of interest, a predictive model will be built using the first 100 observations of the first session as a testing set and the remaining observations as the training set. The models will be compared using the ROC curve and the sensitivity and specificity at the chosen cutoff to determine the most concise model with the best predictive power.

  
# Background

 In the study performed by Steinmetz et al. (2019), experiments were performed on a total of 10 mice over 39 sessions in order to observe how the mice processed stimuli and reacted to it. As stated in the original paper, the purpose was to observe the neurological activities related to choice and activities in the mice. One condition that they presented was that neurological activities correlated with choice must be able to predict chosen actions before they occur. The procedure of this experiment consisted of showing two images of varying contrast levels to the mice, one on each side of the head (left and right). In addition to the contrast levels varying between the left and right sides, the contrast levels could also be the same on each side or no image could be shown at all. The mice were then to select which image had a higher contrast level and turn the wheel of the corresponding side to indicate their decision. If a correct answer was given, the mice were rewarded. In the case of the contrast levels being the same on each side, turning the wheel on either side was rewarded. If there were no images shown, the mice would be rewarded for simply doing nothing for 1.5 seconds. Finally, if the stimulus was present and the mouse did not make any movements for 1.5 seconds after it was presented, then the trial was considered a failure.  

Prior to the experimental sessions, each mouse was trained to achieve high response accuracy and reaction times that were less than 1 second in response to high contrast stimuli. This conditioning continued with no stimuli trials mixed in and then with lower contrast trials. Finally, they were trained in contrast comparison trials. Because each mouse was expected to perform in multiple sessions, any mice that developed extreme health issues were excluded from the analysis. In the original data set, the experimental trials were weighted towards "easier" trials (trials where it was easier to determine which side had higher contrast) in order to maintain motivation and high reward rates throughout the session. Each experimental session consisted of more than 200 trials with a single mouse. Within the collected data set, the contrast levels of the stimuli took on numerical values in {0, 0.25, 0.5, 1}, where 0 indicates the absence of a stimulus. The activity of the neurons in the mice's visual cortex was recorded and made available in the form of spike trains, which is defined as the sequence of neuronal firing timings where the spike indicates that an action potential has been fired. Previous studies have suggested that the firing rate of a neuron is correlated with the strength of the stimulus, with stronger stimuli resulting in higher firing rates (Glazewski,Barth 2014). The trial's success and failure was recorded as either {1} for success or {-1} for failure.  

For this analysis, we focus on the spike trains of neurons in the visual cortex from the onset of the stimuli (t = 0) to 0.4 seconds post-onset. This is because, in the original trial, they observed that the mice would generally turn the wheels during that time period after the stimulus had been presented. Only five sessions (Sessions 1 to 5) with the data from two mice (Cori and Frossman) will be used. 



# Descriptive Analysis  

## Data structure 

---

```{r include=FALSE}
session=list()
for(i in 1:5){
  session[[i]]=readRDS(paste('./Data/session',i,'.rds',sep=''))
}
```

The data came in the form of 5 RDS files that contained the information from the 5 sessions. In each file, the name of the mouse, the date that the experiment was performed on, and 5 other variables could be found. Namely, the variables included are:    

- `feedback_type`: type of the feedback, 1 for success and -1 for failure
- `contrast_left`: contrast of the left stimulus
- `contrast_right`: contrast of the right stimulus
- `time`: centers of the time bins for `spks`  
- `spks`: numbers of spikes of neurons in the visual cortex in time bins defined in `time`  
 

```{r eval=FALSE, include=FALSE}
# Rename eval=TRUE if you want the output to appear in the report.
# Take the 11th trial in Session 1 for example
id=11
session[[1]]$feedback_type[id]
session[[1]]$contrast_left[id]
session[[1]]$contrast_right[id]
length(session[[1]]$time[[id]])
dim(session[[1]]$spks[[id]])

```

```{r}
# Obtain the firing rate 
# averaged over [0,0.4] seconds since stim onsets
# averaged across all neurons 
rate <- list(length(session))

for(j in 1:length(session)){
  
ID=j
t=0.4 # from Background 

n.trials=length(session[[ID]]$spks)
n.neurons=dim(session[[ID]]$spks[[1]])[1]

# Obtain the firing rate 
rate[[j]]=numeric(n.trials)
for(i in 1:n.trials){
  rate[[j]][i]=sum(session[[ID]]$spks[[i]])/n.neurons/t
}
}

#make data for each session
for(i in 1:length(session)){
  
  #index for the session number 
  index <- c(rep(i, dim(session[[i]]$contrast_left)[1]))
   
  #pull out the data of interest
  p =  as.data.frame(cbind(rate[[i]], 
             session[[i]]$contrast_left, 
             session[[i]]$contrast_right,
             session[[i]]$feedback_type, index))
  
  #assign column names
  colnames(p) <- c("fire_rate", "contrast_left", 
                   "contrast_right", "feedback_type", "index")
  
  assign(paste0("session",i),p)
}       

#combining all of the data frames into one
dat <- rbind(session1, session2, session3, session4, session5)

#replacing the -1 with 0
#dat$feedback_type[dat$feedback_type == -1] <- 0

#making the contrasts, feedback, and indexes into factors
dat$index <- as.factor(dat$index)
dat$contrast_left <- as.factor(dat$contrast_left)
dat$contrast_right <- as.factor(dat$contrast_right)
dat$feedback_type <- as.factor(dat$feedback_type)


```

There are no missing values in this data set and no values for the contrasts in the data set that don't conform to the expected values specified in the description. The times, as discussed in the background section, correspond to the times at which the firing data was collected. Spike data was collected 39 times over the 0.4 second interval. 


```{r echo=FALSE, message=FALSE, warning=FALSE}
#Table 1: Numbers of Neurons per session
dims = numeric()

for(j in 1:length(session)){
  dims[j] <- dim(session[[j]]$spks[[1]])[1]
}

counts <- dplyr::summarize(group_by(dat, index, feedback_type),
          n = n())

prop.c <- c(counts[[2,3]] * 100/sum(counts[[1,3]],counts[[2,3]]),
            counts[[4,3]] * 100/sum(counts[[3,3]],counts[[4,3]]),
            counts[[6,3]] * 100/sum(counts[[5,3]],counts[[6,3]]),
            counts[[8,3]] * 100/sum(counts[[7,3]],counts[[8,3]]),
            counts[[10,3]] * 100/sum(counts[[9,3]],counts[[10,3]]))

prop.c <- round(prop.c, 2)

#find proportion of neurons that don't fire per session

n.trials1 <- numeric()

for(j in 1:length(session)){
 n.trials1[j] <-  length(session[[j]]$spks)
}

knitr::kable(cbind(session = c(1:5), 
                   n.trials = n.trials1,
                   n.neurons = dims, 
                   percent.success = prop.c), 
             caption = "Summary of Sessions", 
             align = "ccccc",
             table.attr = "style='width:45%;'")  %>% kable_paper()

```

Table 1 shows a very general summary of the number of trials and neurons observed during each session. Like the number of neurons observed, the numbers of trials (observations per session) also vary, ranging from `r min(length(session[[1]]$spks), length(session[[2]]$spks), length(session[[3]]$spks), length(session[[4]]$spks), length(session[[5]]$spks))` to `r max(length(session[[1]]$spks), length(session[[2]]$spks), length(session[[3]]$spks), length(session[[4]]$spks), length(session[[5]]$spks))`. From the last column of Table 1, it is clear that the proportions of correct to incorrect answers doesn't appear to be too different across the different sessions. It is interesting to note that within each session, there are varying numbers of neurons being observed. The exact numbers of neurons for the 5 sessions are listed in Table 1. The variance in the number of neurons observed in each session has to do with how the original researchers inserted the probe into the brain of the mouse. In each session, the probe was inserted into the same part of the brain (the visual cortex) but the neurons that are actually observed during the session are somewhat random. Because the region of the brain is the same, it is possible for some of the neurons observed in one session to be observed in another session. However, as we do not currently have that information, we cannot know definitively if that is the case. However, an argument can also be made that, despite the probe being inserted into slightly different areas of the brain during each session, the neurons are all from the same side of the visual cortex and should therefore many of these should be neurons of the same general type and or share similar functions. Therefore, they should share similar types of general behaviors regardless of the session. Neurons behave somewhat randomly so it is preferred to have some sort of summary statistic of their behaviors with which to compare them. Thus, the average firing rate of the neurons during a session will be taken as said summary statistic. 

The average firing rates of all of the neurons will be calculated from the number of the spikes recorded across all neurons during the 0.4 second interval. They will be calculated using the equation: $$\small \frac{\text{Sum of Spikes in Trial}}{\text{Number of Neurons} * 0.4}$$ By finding the average firing rate for the trials in each session, the experimental result is condensed into a 1-dimensional observation. This allows us to summarize the observation in the trial as well as compare this value to the other trials in the other session. Additionally, this method of summarizing the neurological data has been used in various other neurological studies. While this method may eliminate any information about the possible clustering of neurons or the firing patterns given certain stimuli, this method creates an easily manageable measure of the activity of the neurons during a certain trial. 

After the data was aggregated, the summary of the data points was checked to make sure that the results made sense. The minimum average firing rate was `r round(min(dat$fire_rate),2)`, which `r ifelse(min(dat$fire_rate > 0), print(", as expected, did not go past zero"), print(",strangely, was less than zero"))`. Also, as the experiment setup from the original study described, there were many instances (above 500 for each) of the left contrast and right contrast taking on the level of 0. From the 5 sessions that we can observe, approximately `r round(nrow(dat[dat$feedback_type == 0,]) * 100 /nrow(dat), 2)`% of the trials were categorized as failures. As seen in Table 1, the session with the highest ratio of failures to successes was session 2.   

In order to understand how the average firing rates varied across different factors, the data will be visualized in a few plots.  


```{r message=FALSE, warning=FALSE, include=FALSE}
#summary statistics of the all data from all sessions 
#summary(dat) 
w <- head(dat, 5)
colnames(w) <- c("firing rate", "left contrast", "right contrast", 
                 "feedback type", "index")
w[,1] <- round(w[,1], 2)

knitr::kable(w,
             caption = "First Few Rows of Aggregated Dataset", 
             align = "cccc",
             table.attr = "style='width:45%;'")  %>% kable_paper()

```


```{r echo=FALSE, fig.cap= "Distribution of Firing Rate Across Sessions", fig.width=9, message=FALSE, warning=FALSE, out.height='35%'}
#graph showing the distributions of all the rates 

a <- ggplot(dat, aes(x = fire_rate)) + 
  geom_density(aes(color = index, fill = index), alpha = 0.3)+
  xlab("Firing Rate") + 
  labs(color = "Session") +
  guides(fill = "none", 
         color = guide_legend(title.hjust = -.5)) + 
  theme(legend.position = c(.95,.95),
        legend.justification = c("right", "top"), 
        legend.box.just = "right",
        legend.key.size = unit(.30, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black")) 



b <- ggplot(dat, aes(x = fire_rate)) + 
  geom_density(alpha = 0.3,
               fill = "grey")+
  xlab("Firing Rate") +
  theme(legend.position = "bottom",
        legend.key.size = unit(.45, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black")) 


f <- ggplot(dat, aes(x = index, y = fire_rate)) + 
  geom_boxplot(aes(color = feedback_type)) + 
  xlab("Session") + 
  ylab("Firing Rate")+ 
  labs(color = "Feedback Type")+
  theme(legend.position = c(.95,.95),
        legend.justification = c("right", "top"), 
        legend.box.just = "right",
        legend.key.size = unit(.30, 'cm'), 
        panel.border = element_blank(),
        axis.line = element_line(color = "black"))

plot_grid(a,b,f, nrow = 1, ncol = 3, labels = 'auto')


w <- t.test(fire_rate ~ feedback_type, session1)



```

Fig.1 shows the distributions of the average firing rates across the different sessions. From Fig.1a, it is clear that there are differences in the firing rates of the neurons across the different sessions. However, it is interesting to note that the curves for sessions 4 and 5 tend to be more similar to each other in both overall shape and value. It’s just that the neurons fired more, on average, during session 4 than in session 5. This can be explained by the two sessions being undergone by the same mouse. However, this explanation does not carry over to the curves for sessions 1,2, and 3, which, despite being undergone by the same mouse, show a large amount of variability. In particular, while the distributions of the average neural activity are relatively similar in sessions 1 and 3, there is a significant decrease in neural activity in session 2.  When observing all of the trials put together, it is clear that the overall distribution, like the distributions by session, appears to be right skewed, an indication of non-normality. Additionally, it appears that much of the information from session 1 is overshadowed by the trends of the firing rates from the other sessions. Thus, it may be more efficient to train the predictive model based on a smaller subset of the data (possibly from a select few sessions) rather than the entire data set. Finally, in Fig.1c, the distributions of the average firing rates per session are plotted based on the feedback type (whether or not the mouse responded correctly). It appears that, in general, the average firing rate appears to be higher in the trials where the mouse responds correctly than the trials where it responds incorrectly. In particular, it seems like, in session 1, the average firing rate of neurons is higher in the successful trials compared to the ones that weren't a success. A hypothesis test at significance level $\alpha = 0.05$ comparing the mean of the average firing rates of the failures and successes in group 1 returns that, with a p-value `r ifelse(w$p.value < 0.05, "smaller than 0.05", "bigger than 0.05")`, the means of the average firing rate between the successful and non-successful trials are significantly different. This is a good indication that there is a sufficient difference in the average firing rates between the outcome types to be able to predict the outcomes of a trial.  


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap= "Feedback Types per Contrast Combination", fig.width=9, message=FALSE, warning=FALSE, out.height='35%'}
c <- ggplot(dat, aes(x = contrast_left,
                     y = contrast_right)) +
  geom_jitter(width = .3, height = 0.3,
              aes(shape = index,
                  color = feedback_type), 
              alpha = 0.75) + 
  xlab("L.Contrast") + 
  ylab("R.Contrast") +
  theme( panel.border = element_blank(),
    axis.line = element_line(color = "black")) 
  

plotly::ggplotly(c)
```

Fig.2 visualizes the distributions of the correct and incorrect responses given the different combinations of the contrasts over the different sessions. It also shows how many of each combination of contrasts were present in each session. As described in the original experimental procedure, the majority of trials had very different contrast levels. Only a small proportion of trials in each session had close or equal contrast levels (excluding the combination {0,0}, which was evaluated using a different set of criteria).  Looking at the plot of the feedback types between the different combinations of contrasts, there are differences in the feedback responses between the contrast levels and across the sessions. For example, in session 1, when the left contrast was {0.25} and the right contrast was {1}, the majority of the responses were labeled as incorrect despite the differences in the contrasts being relatively obvious. A similar result is shown in session 2, when the left contrast is {0} and the right contrast is {1}. Curiously, when the left contrast was {1} and the right contrast was {0}, the correct answer was given the majority of the time in session 2. Despite the difference being relatively great, there are still many incorrect answers being given. These strange patterns of correct and incorrect answers indicates that the left and right contrasts are likely important predictors that should be included in the prediction model.  

```{r message=FALSE, warning=FALSE, include=FALSE}
#density of the firing rates by left contrast and session
d <- ggplot(dat, aes(x = fire_rate)) + 
  geom_density(aes(color = contrast_right, fill = contrast_right), alpha = 0.3)+
  xlab("Firing Rate") + 
  ylab("") +
  labs(color = "R.Contrast") +
  guides(fill = "none", 
         color = guide_legend(title.hjust = -.5)) + 
  theme(legend.box = "vertical",
        legend.title = element_text(size = 10),
        legend.key.size = unit(.45, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black")) +
  facet_wrap(~index)


#density of firing rates by the right contrast and session
e <-  ggplot(dat, aes(x = fire_rate)) + 
  geom_density(aes(color = contrast_left, fill = contrast_left), alpha = 0.3)+
  xlab("Firing Rate") + 
  ylab("") + 
  labs(color = "L.Contrast") +
  guides(fill = "none") + 
  theme(legend.box = "vertical",
        legend.key.size = unit(.45, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    axis.text.y = element_blank()) +
  facet_wrap(~index)

d <- shift_legend2(d)
e <- shift_legend2(e)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap= "Firing Rate by Contrast and Session", fig.topcaption = FALSE, fig.width=9, message=FALSE, warning=FALSE, out.height='35%'}

plot_grid(d,e, nrow = 1, ncol = 2, labels = 'auto')
```

As shown in Fig.3, the densities of the average firing rates of the spikes also tend to differ with the contrast levels and session. Similar to the trends observed in Fig.1, in both the left and right contrast graphs, the curves for sessions 4 and 5 appear to be similar in overall shape while the curves for sessions 1,2, and 3 are notably more different from each other. This is an indication of the average firing rates being more consistent for the mouse Frossman than for the mouse Cori. Also, the firing rates in sessions 1,2,3, and 5 tend to vary a bit with the value of the contrasts, notably when either the left or right contrast takes on the value {1}. When the right contrast is {1} in session 1, there appears to be less neural activity compared to when the left contrast is {1} in the same session. Taken all together, the observations from Fig.1, Fig.2, and Fig.3 point to there being a relationship that is more than just additive between the neural activity and the left and right contrasts.   

```{r eval=FALSE, fig.cap="Main Effect Plots", fig.height=6.5, fig.topcaption=FALSE, fig.width=9, message=FALSE, warning=FALSE, include=FALSE}
#main effects plots and interaction plots

#specifying layout
layout.matrix <- matrix(c(1, 3, 2, 3), nrow = 2, ncol = 2)

graphics::layout(mat = layout.matrix,
       heights = c(1.25, 1.5), # Heights of the two rows
       widths = c(1, 1)) # Widths of the two columns

# Main effect plot for Left Contrast
plotmeans(fire_rate~contrast_left, data=dat,xlab="Left Contrast",ylab="Fire Rate",
          main="Main effect, Left Contrast",cex.lab= 1) 

# Main effect plot for Right Contrast
plotmeans(fire_rate~contrast_right, data=dat,xlab="Right Contrast",ylab="Fire Rate",
          main="Main effect, Right Contrast",cex.lab= 1)

#Interaction plot
interaction.plot(dat$contrast_left, dat$contrast_right, dat$fire_rate
                ,cex.lab= 1,ylab="Fire Rate",xlab='Left Contrast',
                trace.label = "Right Contrast")


par(mfrow = c(1,1))
```

 

## Data Analysis  

### ANOVA Model  

To answer the first question of interest, an analysis using an ANOVA model with mixed effects will be performed. The model will contain the factors for the stimuli from the left and right sides in addition to a random intercept for each session. The general model is as follows:   

$$Y_{ijkl} = \mu_{...} + \alpha_{i} + \beta_{j} + (\alpha \beta)_{ij} + \gamma_{k} + \epsilon_{ijkl}$$ 

:::: {style="display: grid; grid-template-columns: 50% 50% ; grid-column-gap: 5%; "}

::: {}

Where:
    
  - $Y_{ijkl}$ is the average firing rate of a certain trial  
  - $\mu_{...}$ is the population mean  
  - $\alpha_i,\ i = 1,\ldots, 4$ is the fixed effect from the left contrast  
  - $\beta_j,\ j = 1,\ldots, 4$ is the fixed effect from the right contrast  
  - $(\alpha \beta)_{ij}$ is the effects from the interaction of the left and right contrast 
  - $\gamma_{k}$ is the random effect from the session  
  - $\epsilon_{ijkl}$ is the random error term  

:::

::: {}

Constraints/Assumptions:  

  - $\epsilon_{ijk} \sim\  i.i.d\ N(0, \sigma^2)$  
  - $\gamma_{k} \sim\ i.i.d\ N(0, \sigma^2_{\gamma})$  
  - $\epsilon_{ijkl}$ and $\gamma_{k}$ are mutually independent  
  - $\sum \alpha_{i} = 0$  
  - $\sum \beta_{j} =0$  
  - $\sum \sum (\alpha \beta)_{ij} = 0$  

:::

:::: 

    

In this data set, our units of observation are the trails in each session. The strata for this analysis are the different sessions. We include the random effect of the session because there exists a degree of randomness associated with each session. For this analysis, we have randomly selected 5 out of the 39 original sessions to analyze. This random intercept also takes into account any differences there might be between the two mice included in this analysis themselves.  

```{r message=FALSE, warning=FALSE, include=FALSE}
#initial model
sig.level <- c(0.05)

mod1 <- lmer(fire_rate ~ as.factor(contrast_left) + as.factor(contrast_right) + (1 | index),
             data = dat) #reduced model
#summary(mod1)

mod2 <- lmer(fire_rate ~ as.factor(contrast_left) + as.factor(contrast_right) + (1 | index) +
               contrast_left:contrast_right,
             data = dat) #full model


res1 <- anova(mod1, mod2) #returns that the interaction term is significant
#anova(mod1, mod3)

knitr::kable(as.data.frame(mod2@beta))

summary(mod2)

```


Figuring out how neurons respond to stimuli presented on the left and right sides simplifies determining whether or not the left and right stimuli have only additive effects on the neural responses or if they have an interaction effect. This would indicate whether the left and right stimuli are being processed independently or jointly. For this test, $H_0: (\alpha \beta)_{ij} = 0\ \forall i \ and\ j$. In other words, if the null hypothesis is unable to be rejected, then the combination of the contrast level for a trial has no effect on the observed level of neural activity. Under the null hypothesis, the differences in the deviance between the full and reduced model is compared a $\chi^2_{df}$ distribution where $df$ is the degrees of freedom of the variance of the random effects. The test statistic is `r round(res1$Chisq[2], 2)` and the associated p-value is `r round(res1$"Pr(>Chisq)",2)`, indicating significance at the level $\alpha = 0.05$. Therefore, we reject the null hypothesis that the coefficient of the interaction term is zero. Thus, we can conclude that our final model is the same as the general model defined earlier. This result also answers the first question of interest, that the left and right stimuli somehow interact to generate certain levels of neural activity. Additionally, the estimates of the coefficients show that, when either the left or right contrast levels are {1}, the average firing rate tends to increase slightly. These findings are congruent with the conclusions drawn from the observations of Fig.3, where the density curves for the contrast levels {1} tended to be further to the right than some of the other curves.           

### Predictive Model  

```{r message=FALSE, warning=FALSE, include=FALSE}
#predictive model
#split data
dat1 <- dat
levels(dat1$feedback_type)[match("-1",levels(dat1$feedback_type))] <- "0"

train <- dat1[-c(1:100),]
#str(train)

train1 <- dplyr::filter(train, index == 1)
train2 <- dplyr::filter(train, index == 1 | index == 2 | index == 3)
train3 <- dplyr::filter(train, index == 1 | index == 3)

test <- dat1[c(1:100),]

#over all sessions
log.m <- glm(feedback_type ~ fire_rate + contrast_left + contrast_right + index + contrast_left:contrast_right, family = binomial(link='logit'), 
             data = train)

log.gm <- glmer(feedback_type ~ fire_rate + (1 | index) + contrast_left*contrast_right, 
                family = 'binomial',
                control = glmerControl(optimizer = "bobyqa"),
                data = train) #mixed effect logistic model

log.m1 <- glm(feedback_type ~ fire_rate + contrast_left + contrast_right + index, family = binomial(link='logit'), 
             data = train) #remove interaction term

#session 1 only
log.m2 <- glm(feedback_type ~ fire_rate + contrast_left + contrast_right + contrast_left:contrast_right, family = binomial(link='logit'), 
             data = train1)


log.m3 <- glm(feedback_type ~ fire_rate + contrast_left + contrast_right , family = binomial(link='logit'), 
             data = train1) #remove interaction term


#sessions 1 2 and 3

log.m4 <- glm(feedback_type ~ fire_rate + contrast_left + contrast_right + contrast_left:contrast_right, family = binomial(link='logit'), 
             data = train2)

log.m5 <- glm(feedback_type ~ fire_rate + contrast_left + contrast_right , family = binomial(link='logit'), 
             data = train2) #remove interaction term

#sessions 1 and 3 

log.m6 <- glm(feedback_type ~ fire_rate + contrast_left + contrast_right + contrast_left:contrast_right, family = binomial(link='logit'), 
             data = train3)

log.gm1 <- glmer(feedback_type ~ fire_rate + (1 | index) + contrast_left*contrast_right, 
                family = 'binomial',
                control = glmerControl(optimizer = "bobyqa"),
                data = train3) #mixed effect logistic model

log.m7 <- glm(feedback_type ~ fire_rate + contrast_left + contrast_right , family = binomial(link='logit'), 
             data = train3) #remove interaction term


#predicted probabilities
pred.p <-predict(log.m, type = "response") 
pred.p1 <- predict(log.m1, type = "response")
pred.p2 <- predict(log.m2, type = "response")
pred.p3 <- predict(log.m3, type = "response")
pred.p4 <- predict(log.m4, type = "response")
pred.p5 <- predict(log.m5, type = "response")
pred.p6 <- predict(log.m6, type = "response")
pred.p7 <- predict(log.m7, type = "response")
pred.p8 <- predict(log.gm, type = "response")
pred.p9 <- predict(log.gm1, type = "response")

ROC <- pROC::roc(train$feedback_type ~ pred.p)
ROC1 <- pROC::roc(train$feedback_type ~ pred.p1)
ROC2 <- pROC::roc(train1$feedback_type ~ pred.p2)
ROC3 <- pROC::roc(train1$feedback_type ~ pred.p3)
ROC4 <- pROC::roc(train2$feedback_type ~ pred.p4)
ROC5 <- pROC::roc(train2$feedback_type ~ pred.p5)
ROC6 <- pROC::roc(train3$feedback_type ~ pred.p6)
ROC7 <- pROC::roc(train3$feedback_type ~ pred.p7)
ROC8 <- pROC::roc(train$feedback_type ~ pred.p8)
ROC9 <- pROC::roc(train3$feedback_type ~ pred.p9)

plot(ROC)
plot(ROC1, add = TRUE, col = 'red')
plot(ROC2, add = TRUE, col = 'blue')
plot(ROC3, add = TRUE, col = 'orange')
plot(ROC4, add = TRUE, col = 'purple')
plot(ROC5, add = TRUE, col = 'green')
plot(ROC6, add = TRUE, col = 'grey')
plot(ROC7, add = TRUE, col = 'brown')
plot(ROC8, add = TRUE, col = 'turquoise')
plot(ROC9, add = TRUE, col = 'darkgreen')


pROC::auc(train1$feedback_type ~ pred.p3)

#testing different cutoffs for this model
## function for easily assigning the predicted values 
log.pred <- function(mod, data, pos = 1, neg = 0, cut = 0.5){
  prob <- predict(mod, newdata = data, type = "response")
  pred <- ifelse(prob > cut, pos, neg)
  
  #classification error rate
  actual <- data[,"feedback_type"]
  c.err <- mean(actual != pred)
  
  #confusion matrix
  tab <- table(pred, actual)
  
  con.mat <- confusionMatrix(tab)
  accuracy <- con.mat$overall["Accuracy"]
  sensitivity <- con.mat$byClass["Sensitivity"]
  specificity <- con.mat$byClass["Specificity"]
  
  return(list(c.err, tab, accuracy, sensitivity, specificity))
}

log.pred(log.m3, data = test, cut = 0.69) #best
log.pred(log.m, data = test, cut = 0.69)
log.pred(log.gm1, data = test, cut = 0.69)
```

In order to answer the second question of interest, a predictive model was built. Because the response that we are interested in predicting is a binary factor, a log odds model was the obvious choice. The log odds model, for this data set, will take the form:  

$$ln(\frac{P}{1-P}) = \beta_0 + \beta_{1i}X_{1i} +\sum_{j = 1}^{4}(\beta_{2j}X_{2j} + \beta_{3j}X_{3j}) + \epsilon_{i}$$
Where:  
$\beta_1$ and $X_1$ correspond to the average firing rates from the first session and $\beta_2$, $X_2$, $\beta_3$, and $X_3$ correspond to the effects of the left and right contrasts respectively. $\epsilon_i$ is the error term. The designated cutoff level for this model was p = .69.  

When deciding on how to build the predictive model, an issue arose regarding the data with which the model will be trained on. This model will be validated using the first 100 observations of the first session, meaning that all of these observations come from a single session and a single mouse. As shown in the Fig.1, the firing rates are very different with respect to the session and the mouse. In Frossman's trials, the average firing rate of his neurons is relatively lower than the firing rates of Cori's neurons. Given these observations, models fitted with different training sets were also compared to each other. The options for the training data were the full data set, session 1 only, sessions 1, 2,and 3, and only sessions 1 and 3. To compare the models and their performance, ROC curves were plotted to visualize the relationships between sensitivity and specificity for all of the models. Of all of the models visualizes, the model trained only with session 1 data had the largest area under the curve (AUC). With p = 0.69, the accuracy came out to be 72% and the sensitivity and specificity were 0.42 and 0.82 respectively. Out of all of the models that were tested, this one performed relatively better. The other two similarly performing models were a model that had the same general formula but was trained on data from sessions 1 and 3 and a model that was also trained on sessions 1 and 3 but contained a random effect for the session. The first of these two models had an accuracy of 68%, a sensitivity of .53 and specificity of .73. The second model had an accuracy of 70% and had .38 for the sensitivity and .81 for the specificity. Compared to these two models, the final model chosen had better accuracy and specificity. The confusion matrix for the final model is shown below in Table 2.  

```{r echo=FALSE}
tab.c <- data.frame(cbind(c(11,15), c(13,61))) 
rownames(tab.c) <- c("Pred.0", "Pred.1") 
colnames(tab.c) <- c("Actual.0", "Actual.1")

knitr::kable(tab.c, 
             caption = "Confusion Matrix of Prediction Model", 
             align = "cc",
             table.attr = "style='width:45%;'")  %>% kable_paper()

```


## Sensitivity Analysis  

### Analysis of residuals for ANOVA Model   

```{r echo=FALSE, fig.height=4, fig.width=8.5, message=FALSE, warning=FALSE, fig.cap = "Residual Plot of Mixed Effect Model"}
#residual plot over time of the model

resid <- residuals(mod2)

trial <- c(seq(1,length(session[[1]]$spks)),
           seq(1,length(session[[2]]$spks)),
           seq(1,length(session[[3]]$spks)),
           seq(1,length(session[[4]]$spks)),
           seq(1,length(session[[5]]$spks)))

index <-  c(rep(1, length(session[[1]]$spks)),
            rep(2, length(session[[2]]$spks)),
            rep(3, length(session[[3]]$spks)),
            rep(4, length(session[[4]]$spks)),
            rep(5, length(session[[5]]$spks)))

outcome <-c(session1$feedback_type,
                 session2$feedback_type, 
                 session3$feedback_type, 
                 session4$feedback_type, 
                 session5$feedback_type)



tab.resid <- as.data.frame(cbind(resid, trial , 
                                 outcome,
                                 index = as.factor(index)))

tab.resid$outcome <- as.factor(outcome)

ggplot(tab.resid, aes(x = trial, y = resid)) + 
  geom_point(aes(color = outcome), 
             alpha = 0.7) + 
  geom_smooth(se = T, method = 'lm',
              color = "orange") + 
  facet_wrap(~index, ncol = 5) + 
  theme(axis.text.x = element_blank(),
        legend.box = "vertical",
        legend.key.size = unit(.45, 'cm'), 
        legend.position = c(.95,.95),
        legend.justification = c("right", "top"), 
        legend.box.just = "right",
        panel.border = element_blank())

#plot(resid ~ fitted(mod2))

```

In Fig.4, the plots of the residuals over time for the mixed effects ANOVA model are shown. It is clear that there is some sort of trend in each session. Overall, there is a negative correlation between the residuals and the number of trials. This corresponds to the average firing rate decreasing over the course of the trials. One possible explanation is that, over the course of each session, the mouse becomes tired and so the average firing rate of its neurons decreases. Another possible explanation could be that the mouse's neurons are becoming more adapted to or accustomed to certain patterns so the neuronal connections that are required to answer are become strengthened while other connections are being eliminated, leading to overall lower levels of activity. Because the mice have already been trained on the procedures, it's possible that these observations are simply the pruning of unnecessary mental processes for this particular task. However, there isn't really any conclusive evidence pointing towards either of these hypotheses. There doesn't appear to be a trend of increasing density of "successful" trials as the sessions go on so it's hard to say definitively if the mouse is improving or performing worse with time. 

### Comparing the Mean vs. Other Test Statistics  

```{r echo=FALSE, fig.cap="Number of Spks Per Neuron", fig.height=3, fig.topcaption=FALSE, fig.width=9, message=FALSE, warning=FALSE, paged.print=TRUE}
for(j in 1:length(session)){
  
x <- session[[j]]$spks
augh <- matrix(rep(0, length(x) * dim(x[[1]])[1]), ncol = dim(x[[1]])[1], nrow = length(x))

for(i in 1:length(x)){ #number of trials
  
  augh[i,] <- apply(x[[i]], 1, sum) 
  
}
assign(paste0("r", j), melt(augh))
}
t <- 0.4 

#getting rates for each neuron

#r1$value <- r1$value/t
#r2$value <- r2$value/t
#r3$value <- r3$value/t
#r4$value <- r4$value/t
#r5$value <- r5$value/t


r1[,"index"] <- rep(1, nrow(r1))

r2[,"index"] <- rep(2, nrow(r2))

r3[,"index"] <- rep(3, nrow(r3))

r4[,"index"] <- rep(4, nrow(r4))

r5[,"index"] <- rep(5, nrow(r5))

r <- rbind(r1, r2, r3, r4, r5)

ggplot(r, aes(x = value, y = ..density.., group = as.factor(index))) + 
  geom_histogram(bins = 20) + 
  facet_grid(~index) + 
  ylim(0,0.059) + 
  xlab("Total Spks of Neuron in a Trial")

```

```{r message=FALSE, warning=FALSE, include=FALSE}
#maximum as test statistic

rate1 <- list(length(session))

for(j in 1:length(session)){
  
  ID=j
  t=0.4 # from Background 
  
  n.trials=length(session[[ID]]$spks)
  n.neurons=dim(session[[ID]]$spks[[1]])[1]
  
  # Obtain the firing rate 
  rate1[[j]]=numeric(n.trials)
  for(i in 1:n.trials){
    rate1[[j]][i]=max(session[[ID]]$spks[[i]])/t
  }
}

#make data for each session
for(i in 1:length(session)){
  
  #index for the session number 
  index <- c(rep(i, dim(session[[i]]$contrast_left)[1]))
  
  #pull out the data of interest
  p =  as.data.frame(cbind(rate1[[i]], 
                           session[[i]]$contrast_left, 
                           session[[i]]$contrast_right,
                           session[[i]]$feedback_type, index))
  
  #assign column names
  colnames(p) <- c("fire_rate", "contrast_left", 
                   "contrast_right", "feedback_type", "index")
  
  assign(paste0("session.",i),p)
}       

#combining all of the data frames into one
dat2 <- rbind(session.1, session.2, session.3, session.4, session.5)

#replacing the -1 with 0
#dat$feedback_type[dat$feedback_type == -1] <- 0

#making the contrasts, feedback, and indexes into factors
dat2$index <- as.factor(dat$index)
dat2$contrast_left <- as.factor(dat$contrast_left)
dat2$contrast_right <- as.factor(dat$contrast_right)
dat2$feedback_type <- as.factor(dat$feedback_type)

summary(dat2)

#remake fig.1c with the maximum data
ggplot(dat2, aes(x = index, y = fire_rate)) + 
  geom_boxplot(aes(color = feedback_type)) + 
  xlab("Session") + 
  ylab("Firing Rate")+ 
  labs(color = "Feedback Type")+
  theme(legend.position = c(.95,.95),
        legend.justification = c("right", "top"), 
        legend.box.just = "right",
        legend.key.size = unit(.30, 'cm'), 
        panel.border = element_blank(),
        axis.line = element_line(color = "black"))
```


When considering other test statistics, the median of the firing rates is a genuine consideration. However, looking at the distributions of the total number of times a neuron was fired during a trial, you'd find that the majority of the neurons don't fire at all during these trials. Therefore, taking the median of these firing rates would return a value that's very close to zero and thus, very uninformative for our purposes. The maximum and minimum can also be ruled out using similar reasons. For obvious reasons, the minimum (which is in all cases 0) would be uninformative. The maximum, on the other hand, would be ruled out due to instances of strong neural activity, sometimes resulting in a total of over 30 spikes in a single trial from a single neuron. When actually testing the maximum firing rate across the different trails of the different sessions, it became clear that any trends that were observed while using the mean firing rate vanished when using the maximum. Also, using the maximum and or minimum would make the analysis difficult to interpret. The same can be said for a statistic like the standard deviation.  

### Comparison Between Fixed Effects and Mixed Effects Models  

```{r message=FALSE, warning=FALSE, include=FALSE}
#fixed model

fix <- lm(fire_rate ~ as.factor(contrast_left) + as.factor(contrast_right) + index +
               contrast_left:contrast_right,
             data = dat)

#likelihood ratio test
lrtest(mod2, fix)

```

In order to verify the necessity of the random effects of session within the model, a likelihood ratio test will be performed between a model with fixed effects for session and the mixed effect model. Because the likelihood-ratio test between the model with mixed effects and the model with fixed effects returns a p-value much less than 0.05, we reject the null hypothesis that the fixed effect model and mixed effect model fit the data equally well. In other words, the mixed effect model fits our observed data better than the fixed effect model.

### Comparison With Results from Cluster Analysis  

As mentioned in an earlier section, using the mean as a test statistic is a relatively crude method that overlooks the firing patterns of certain neurons and that effect on the overall firing rates. From Fig.5 above, it is worth noting that not all of the neurons behave the same way when the stimuli are presented. The neurons can be broken down into three groups: neurons that don't fire much when the stimuli are presented, neurons that fire a moderate amount, and neurons that fire a lot. Therefore, it should theoretically be possible to divide the neurons into different "types" based on their reactions to stimuli. In this attempt at clustering the data, each individual trial was condensed into the sum of the number of spks per neuron. Then, these values were averaged by the trial type (the specific combinations of left and right contrasts for that trial). This way, each neuron from each session is compared to its average firing rate by the trial type. The K-means clustering algorithm with k = 3 was used to cluster the neurons. Each neuron was assigned to an index (1, 2, or 3) designating the cluster. In this case, 1 corresponds to "medium" activity, 2 corresponds "high" activity, and 3 corresponds to "low" activity. Session 5 had no neurons with "high" activity levels.   

```{r echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
set.seed(1)

#first try to get the files sorted by the neurons

# first get the sums of the firings per neuron in each session

for(j in 1:length(session)){
  
  x <- session[[j]]$spks
  augh <- matrix(rep(0, length(x) * dim(x[[1]])[1]), ncol = dim(x[[1]])[1], nrow = length(x))
  
  for(i in 1:length(x)){ #number of trials
    
    augh[i,] <- apply(x[[i]], 1, sum) 
    
  }
  assign(paste0("r", j), reshape2::melt(augh))
}

#naming the columns 
colnames(r1) <- c("trial", "neuron", "num_spks")
colnames(r2) <- c("trial", "neuron", "num_spks")
colnames(r3) <- c("trial", "neuron", "num_spks")
colnames(r4) <- c("trial", "neuron", "num_spks")
colnames(r5) <- c("trial", "neuron", "num_spks")

#just getting the totals per neuron
b1 <- summarise(group_by(r1, neuron), 
                spks = sum(num_spks))
b2 <- summarise(group_by(r2, neuron), 
                spks = sum(num_spks))
b3 <- summarise(group_by(r3, neuron), 
                spks = sum(num_spks))
b4 <- summarise(group_by(r4, neuron), 
                spks = sum(num_spks))
b5 <- summarise(group_by(r5, neuron), 
                spks = sum(num_spks))

#putting all the totals into one data frame


#get trial types for every trial

for(i in 1:length(session)){
  
  #index for the session number 
  index <- c(rep(i, dim(session[[i]]$contrast_left)[1]))
  
  #pull out the data of interest
  v <- cbind(session[[i]]$contrast_left, 
                     session[[i]]$contrast_right,
                     session[[i]]$feedback_type, 
                     index)
  
  v <- as.data.frame(v)
  
  #assign column names
  colnames(v) <- c("contrast_left", 
                   "contrast_right", 
                   "feedback_type", 
                   "index")
  
  assign(paste0("s",i), v)
}      

#finding pairs of trial types
for(i in 1:length(session)){
  
  type <- numeric()
  
  for(j in 1:length(session[[i]]$contrast_left)){
    
    #assigning a number for the 16 trial types
  if(get(paste0("s",i))[j,1] == 0 & get(paste0("s",i))[j,2] == 0 ){
    type[j] <- 1
  } else if(get(paste0("s",i))[j,1] == 0.25 & get(paste0("s",i))[j,2] == 0 ){
    type[j] <- 2
  } else if(get(paste0("s",i))[j,1] == 0.5 & get(paste0("s",i))[j,2] == 0 ){
    type[j] <- 3
  } else if(get(paste0("s",i))[j,1] == 1 & get(paste0("s",i))[j,2] == 0 ){
    type[j] <- 4
  } else if(get(paste0("s",i))[j,1] == 0 & get(paste0("s",i))[j,2] == 0.25 ){
    type[j] <- 5
  } else if(get(paste0("s",i))[j,1] == 0.25 & get(paste0("s",i))[j,2] == 0.25 ){
    type[j] <- 6
  } else if(get(paste0("s",i))[j,1] == 0.5 & get(paste0("s",i))[j,2] == 0.25 ){
    type[j] <- 7
  } else if(get(paste0("s",i))[j,1] == 1 & get(paste0("s",i))[j,2] == 0.25 ){
    type[j] <- 8
  } else if(get(paste0("s",i))[j,1] == 0 & get(paste0("s",i))[j,2] == 0.5 ){
    type[j] <- 9
  } else if(get(paste0("s",i))[j,1] == 0.25 & get(paste0("s",i))[j,2] == 0.5 ){
    type[j] <- 10
  } else if(get(paste0("s",i))[j,1] == 0.5 & get(paste0("s",i))[j,2] == 0.5 ){
    type[j] <- 11
  } else if(get(paste0("s",i))[j,1] == 1 & get(paste0("s",i))[j,2] == 0.5 ){
    type[j] <- 12
  } else if(get(paste0("s",i))[j,1] == 0 & get(paste0("s",i))[j,2] == 1 ){
    type[j] <- 13
  } else if(get(paste0("s",i))[j,1] == 0.25 & get(paste0("s",i))[j,2] == 1 ){
    type[j] <- 14
  } else if(get(paste0("s",i))[j,1] == 0.5 & get(paste0("s",i))[j,2] == 1 ){
    type[j] <- 15
  } else if(get(paste0("s",i))[j,1] == 1 & get(paste0("s",i))[j,2] == 1 ){
    type[j] <- 16
  } else {}
  }
  assign(paste0("t",i), type)
}

#matching the types to the neuron data
r1["type"] <- c(rep(t1, dim(session[[1]]$spks[[1]])[1]))
r2["type"] <- c(rep(t2, dim(session[[2]]$spks[[1]])[1]))
r3["type"] <- c(rep(t3, dim(session[[3]]$spks[[1]])[1]))
r4["type"] <- c(rep(t4, dim(session[[4]]$spks[[1]])[1]))
r5["type"] <- c(rep(t5, dim(session[[5]]$spks[[1]])[1]))

t = 0.4

#find avg spike rate by trial type
w1 <- summarise(group_by(r1, type, neuron),
                rate = mean(num_spks)/t)
w2 <- summarise(group_by(r2, type, neuron),
                rate = mean(num_spks)/t)
w3 <- summarise(group_by(r3, type, neuron),
                rate = mean(num_spks)/t)
w4 <- summarise(group_by(r4, type, neuron),
                rate = mean(num_spks)/t)
w5 <- summarise(group_by(r5, type, neuron),
                rate = mean(num_spks)/t)

#make this into a matrix/array
tab1 <- matrix(w1$rate, 
               nrow = dim(session[[1]]$spks[[1]])[1],
               ncol = 16)
tab2 <- matrix(w2$rate, 
               nrow = dim(session[[2]]$spks[[1]])[1],
               ncol = 16)
tab3 <- matrix(w3$rate, 
               nrow = dim(session[[3]]$spks[[1]])[1],
               ncol = 16)
tab4 <- matrix(w4$rate, 
               nrow = dim(session[[4]]$spks[[1]])[1],
               ncol = 16)
tab5 <- matrix(w5$rate, 
               nrow = dim(session[[5]]$spks[[1]])[1],
               ncol = 16)

tab <- rbind(tab1, tab2, tab3, tab4, tab5)
colnames(tab) <- c("(0,0)", "(0.25, 0)", "(0.5, 0)", "(1,0)",
                   "(0, 0.25)", "(0.25, 0.25)", "(0.5, 025)", "(1, 0.25)",
                   "(0, 0.5)", "(0.25, 0.5)", "(0.5, 0.5)", "(1, 0.5)",
                   "(0, 1)", "(0.25, 1)", "(0.5, 1)", "(1,1)"
)

#K-means cluster analysis
clusters <- kmeans(tab, 3)

index <-  c(rep(1, dim(session[[1]]$spks[[1]])[1]),
            rep(2, dim(session[[2]]$spks[[1]])[1]),
            rep(3, dim(session[[3]]$spks[[1]])[1]),
            rep(4, dim(session[[4]]$spks[[1]])[1]),
            rep(5, dim(session[[5]]$spks[[1]])[1]))

#adding cluster assignment to the session numbers
clusters1 <- as.data.frame(cbind(clusters = clusters$cluster,
                           index))

c1 <- filter(clusters1, index == 1)
c2 <- filter(clusters1, index == 2)
c3 <- filter(clusters1, index == 3)
c4 <- filter(clusters1, index == 4)
c5 <- filter(clusters1, index == 5)

#sorting data of sums by the trial number and neuron numbers
r1 <- r1[
    order( r1[,1], r1[,2] ),
  ]
r2 <- r2[
  order( r2[,1], r2[,2] ),
]
r3 <- r3[
  order( r3[,1], r3[,2] ),
]
r4 <- r4[
  order( r4[,1], r4[,2] ),
]
r5 <- r5[
  order( r5[,1], r5[,2] ),
]

#append cluster assignment 

r1["clust"] <- c(rep(c1$clusters, length(session[[1]]$spks)))
r2["clust"] <- c(rep(c2$clusters, length(session[[2]]$spks)))
r3["clust"] <- c(rep(c3$clusters, length(session[[3]]$spks)))
r4["clust"] <- c(rep(c4$clusters, length(session[[4]]$spks)))
r5["clust"] <- c(rep(c5$clusters, length(session[[5]]$spks)))

#splitting the data by cluster

#cluster 1
r1_c1 <- as.data.frame(cbind(summarise(group_by(filter(r1, clust == 1), trial),
                  fire_rate = mean(num_spks)/t),
                  contrast_left = session[[1]]$contrast_left, 
                  contrast_right = session[[1]]$contrast_right,
                  feedback_type = session[[1]]$feedback_type, 
                  index = rep(1, length(session[[1]]$spks))))
r2_c1 <- as.data.frame(cbind(summarise(group_by(filter(r2, clust == 1), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[2]]$contrast_left, 
                             contrast_right = session[[2]]$contrast_right,
                             feedback_type = session[[2]]$feedback_type, 
                       index = rep(2, length(session[[2]]$spks))))
r3_c1 <- as.data.frame(cbind(summarise(group_by(filter(r3, clust == 1), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[3]]$contrast_left, 
                             contrast_right = session[[3]]$contrast_right,
                             feedback_type = session[[3]]$feedback_type, 
                       index = rep(3, length(session[[3]]$spks)))) 
r4_c1 <- as.data.frame(cbind(summarise(group_by(filter(r4, clust == 1), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[4]]$contrast_left, 
                             contrast_right = session[[4]]$contrast_right,
                             feedback_type = session[[4]]$feedback_type, 
                       index = rep(4, length(session[[4]]$spks))))
r5_c1 <- as.data.frame(cbind(summarise(group_by(filter(r5, clust == 1), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[5]]$contrast_left, 
                             contrast_right = session[[5]]$contrast_right,
                             feedback_type = session[[5]]$feedback_type, 
                       index = rep(5, length(session[[5]]$spks))))

clust1 <- rbind(r1_c1, r2_c1, r3_c1,
                r4_c1, r5_c1)
#making the contrasts, feedback, and indexes into factors
clust1$index <- as.factor(clust1$index)
clust1$contrast_left <- as.factor(clust1$contrast_left)
clust1$contrast_right <- as.factor(clust1$contrast_right)
clust1$feedback_type <- as.factor(clust1$feedback_type)

#cluster 2
r1_c2 <- as.data.frame(cbind(summarise(group_by(filter(r1, clust == 2), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[1]]$contrast_left, 
                             contrast_right = session[[1]]$contrast_right,
                             feedback_type = session[[1]]$feedback_type, 
                             index = rep(1, length(session[[1]]$spks))))
r2_c2 <- as.data.frame(cbind(summarise(group_by(filter(r2, clust == 2), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[2]]$contrast_left, 
                             contrast_right = session[[2]]$contrast_right,
                             feedback_type = session[[2]]$feedback_type, 
                             index = rep(2, length(session[[2]]$spks))))
r3_c2 <- as.data.frame(cbind(summarise(group_by(filter(r3, clust == 2), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[3]]$contrast_left, 
                             contrast_right = session[[3]]$contrast_right,
                             feedback_type = session[[3]]$feedback_type, 
                             index = rep(3, length(session[[3]]$spks)))) 
r4_c2 <- as.data.frame(cbind(summarise(group_by(filter(r4, clust == 2), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[4]]$contrast_left, 
                             contrast_right = session[[4]]$contrast_right,
                             feedback_type = session[[4]]$feedback_type, 
                             index = rep(4, length(session[[4]]$spks))))

#r5_c2 <- as.data.frame(cbind(summarise(group_by(filter(r5, clust == 2), trial),
                                    #   fire_rate = mean(num_spks)/t),
                            # contrast_left = session[[5]]$contrast_left, 
                            # contrast_right = session[[5]]$contrast_right,
                            # feedback_type = session[[5]]$feedback_type, 
                             #index = rep(5, length(session[[5]]$spks))))#no clust 2

clust2 <- rbind(r1_c2, r2_c2, r3_c2,
                r4_c2)
#making the contrasts, feedback, and indexes into factors
clust2$index <- as.factor(clust2$index)
clust2$contrast_left <- as.factor(clust2$contrast_left)
clust2$contrast_right <- as.factor(clust2$contrast_right)
clust2$feedback_type <- as.factor(clust2$feedback_type)


#cluster 3
r1_c3 <- as.data.frame(cbind(summarise(group_by(filter(r1, clust == 3), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[1]]$contrast_left, 
                             contrast_right = session[[1]]$contrast_right,
                             feedback_type = session[[1]]$feedback_type, 
                             index = rep(1, length(session[[1]]$spks))))
r2_c3 <- as.data.frame(cbind(summarise(group_by(filter(r2, clust == 3), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[2]]$contrast_left, 
                             contrast_right = session[[2]]$contrast_right,
                             feedback_type = session[[2]]$feedback_type, 
                             index = rep(2, length(session[[2]]$spks))))
r3_c3 <- as.data.frame(cbind(summarise(group_by(filter(r3, clust == 3), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[3]]$contrast_left, 
                             contrast_right = session[[3]]$contrast_right,
                             feedback_type = session[[3]]$feedback_type, 
                             index = rep(3, length(session[[3]]$spks)))) 
r4_c3 <- as.data.frame(cbind(summarise(group_by(filter(r4, clust == 3), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[4]]$contrast_left, 
                             contrast_right = session[[4]]$contrast_right,
                             feedback_type = session[[4]]$feedback_type, 
                             index = rep(4, length(session[[4]]$spks))))
r5_c3 <- as.data.frame(cbind(summarise(group_by(filter(r5, clust == 3), trial),
                                       fire_rate = mean(num_spks)/t),
                             contrast_left = session[[5]]$contrast_left, 
                             contrast_right = session[[5]]$contrast_right,
                             feedback_type = session[[5]]$feedback_type, 
                             index = rep(5,length(session[[5]]$spks))))

clust3 <- rbind(r1_c3, r2_c3, r3_c3,
                r4_c3, r5_c3)
#making the contrasts, feedback, and indexes into factors
clust3$index <- as.factor(clust3$index)
clust3$contrast_left <- as.factor(clust3$contrast_left)
clust3$contrast_right <- as.factor(clust3$contrast_right)
clust3$feedback_type <- as.factor(clust3$feedback_type)

#fit anova models
#cluster 1
mod1 <- lmer(fire_rate ~ as.factor(contrast_left) + as.factor(contrast_right) + (1 | index),
             data = clust1) #reduced model

mod2 <- lmer(fire_rate ~ as.factor(contrast_left) + as.factor(contrast_right) + (1 | index) +
               contrast_left:contrast_right,
             data = clust1) #full model

res2 <- anova(mod1, mod2)

#cluster 2
mod3 <- lmer(fire_rate ~ as.factor(contrast_left) + as.factor(contrast_right) + (1 | index),
             data = clust2) #reduced model

mod4 <- lmer(fire_rate ~ as.factor(contrast_left) + as.factor(contrast_right) + (1 | index) +
               contrast_left:contrast_right,
             data = clust2) #full model

res3 <- anova(mod3, mod4)

#cluster 3
mod5 <- lmer(fire_rate ~ as.factor(contrast_left) + as.factor(contrast_right) + (1 | index),
             data = clust3) #reduced model

mod6 <- lmer(fire_rate ~ as.factor(contrast_left) + as.factor(contrast_right) + (1 | index) +
               contrast_left:contrast_right,
             data = clust3) #full model

res4 <- anova(mod5, mod6)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
#density of the firing rates by left contrast and session
f <- ggplot(clust1, aes(x = fire_rate)) + 
  geom_density(aes(color = contrast_right, fill = contrast_right), alpha = 0.3)+
  xlab("Firing Rate") + 
  ylab("") +
  labs(color = "R.Contrast") +
  guides(fill = "none", 
         color = guide_legend(title.hjust = -.5)) + 
  theme(legend.box = "vertical",
        legend.title = element_text(size = 10),
        legend.key.size = unit(.45, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black")) +
  facet_wrap(~index)


#density of firing rates by the right contrast and session
g <-  ggplot(clust1, aes(x = fire_rate)) + 
  geom_density(aes(color = contrast_left, fill = contrast_left), alpha = 0.3)+
  xlab("Firing Rate") + 
  ylab("") + 
  labs(color = "L.Contrast") +
  guides(fill = "none") + 
  theme(legend.box = "vertical",
        legend.key.size = unit(.45, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    axis.text.y = element_blank()) +
  facet_wrap(~index)

f <- shift_legend2(f)
g <- shift_legend2(g)
```

```{r fig.cap="Firing Rate by Contrast and Session For Cluster 1", fig.topcaption=FALSE, fig.width=9, message=FALSE, warning=FALSE, include=FALSE, out.height='35%'}

plot_grid(f,g, nrow = 1, ncol = 2, labels = 'auto')
```

```{r message=FALSE, warning=FALSE, include=FALSE}
#density of the firing rates by left contrast and session
h <- ggplot(clust2, aes(x = fire_rate)) + 
  geom_density(aes(color = contrast_right, fill = contrast_right), alpha = 0.3)+
  xlab("Firing Rate") + 
  ylab("") +
  labs(color = "R.Contrast") +
  guides(fill = "none", 
         color = guide_legend(title.hjust = -.5)) + 
  theme(legend.box = "vertical",
        legend.title = element_text(size = 10),
        legend.key.size = unit(.45, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black")) +
  facet_wrap(~index)


#density of firing rates by the right contrast and session
i <-  ggplot(clust2, aes(x = fire_rate)) + 
  geom_density(aes(color = contrast_left, fill = contrast_left), alpha = 0.3)+
  xlab("Firing Rate") + 
  ylab("") + 
  labs(color = "L.Contrast") +
  guides(fill = "none") + 
  theme(legend.box = "vertical",
        legend.key.size = unit(.45, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    axis.text.y = element_blank()) +
  facet_wrap(~index)

```

```{r fig.cap="Firing Rate by Contrast and Session For Cluster 2", fig.topcaption=FALSE, fig.width=9, message=FALSE, warning=FALSE, include=FALSE, out.height='35%'}

plot_grid(h,i, nrow = 1, ncol = 2, labels = 'auto')
```

```{r message=FALSE, warning=FALSE, include=FALSE}
#density of the firing rates by left contrast and session
j <- ggplot(clust3, aes(x = fire_rate)) + 
  geom_density(aes(color = contrast_right, fill = contrast_right), alpha = 0.3)+
  xlab("Firing Rate") + 
  ylab("") +
  labs(color = "R.Contrast") +
  guides(fill = "none", 
         color = guide_legend(title.hjust = -.5)) + 
  theme(legend.box = "vertical",
        legend.title = element_text(size = 10),
        legend.key.size = unit(.45, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black")) +
  facet_wrap(~index)


#density of firing rates by the right contrast and session
k <-  ggplot(clust3, aes(x = fire_rate)) + 
  geom_density(aes(color = contrast_left, fill = contrast_left), alpha = 0.3)+
  xlab("Firing Rate") + 
  ylab("") + 
  labs(color = "L.Contrast") +
  guides(fill = "none") + 
  theme(legend.box = "vertical",
        legend.key.size = unit(.45, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    axis.text.y = element_blank()) +
  facet_wrap(~index)

j <- shift_legend2(j)
k <- shift_legend2(k)
```

```{r fig.cap="Firing Rate by Contrast and Session For Cluster 3", fig.topcaption=FALSE, fig.width=9, message=FALSE, warning=FALSE, include=FALSE, out.height='35%'}

plot_grid(j,k, nrow = 1, ncol = 2, labels = 'auto')
```

```{r echo=FALSE, fig.cap= "Distribution of Firing Rate Across Sessions and Clusters", fig.width=9, message=FALSE, warning=FALSE, out.height='35%'}
#graph showing the distributions of all the rates 

a <- ggplot(clust1, aes(x = fire_rate)) + 
  geom_density(aes(color = index, fill = index), alpha = 0.3)+
  xlab("Firing Rate") + 
  labs(color = "Session", 
       title = "Cluster 1") +
  guides(fill = "none", 
         color = guide_legend(title.hjust = -.5)) + 
  theme(legend.position = c(.95,.95),
        legend.justification = c("right", "top"), 
        legend.box.just = "right",
        legend.key.size = unit(.30, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black")) 



b <-  ggplot(clust2, aes(x = fire_rate)) + 
  geom_density(aes(color = index, fill = index), alpha = 0.3)+
  xlab("Firing Rate") + 
  labs(color = "Session",
       title = "Cluster 2") +
  guides(fill = "none", 
         color = guide_legend(title.hjust = -.5)) + 
  theme(legend.position = c(.95,.95),
        legend.justification = c("right", "top"), 
        legend.box.just = "right",
        legend.key.size = unit(.30, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black")) 


f <-  ggplot(clust3, aes(x = fire_rate)) + 
  geom_density(aes(color = index, fill = index), alpha = 0.3)+
  xlab("Firing Rate") + 
  labs(color = "Session",
       title = "Cluster 3") +
  guides(fill = "none", 
         color = guide_legend(title.hjust = -.5)) + 
  theme(legend.position = c(.95,.95),
        legend.justification = c("right", "top"), 
        legend.box.just = "right",
        legend.key.size = unit(.30, 'cm'), 
        panel.border = element_blank(),
    axis.line = element_line(color = "black")) 

plot_grid(a,b,f, nrow = 1, ncol = 3, labels = 'auto')

```

As seen in Fig.6, the distributions of the firing rates does in fact differ by cluster. Session 5 has no neurons classified as "high" activity and session 4 does not have many neurons of that category either. This could be due to the number of neurons observed as session 5 only has 99 and session 4 has 120. In cluster 1, the distributions for sessions 4 and 5 closely resemble each other, a trend also seen in the overall means analysis. The significance of the interaction term was also tested for the different clusters. The ANOVA model was of the same form as the model introduced earlier, just run on different data sets split by cluster. In all three, when the full and reduced models were compared, the interactions between the left and right contrasts were deemed to not be significant at the significance level $\alpha = 0.05$, which is very different from the results of the initial analysis, which concluded that the interaction was a significant influence on the average firing rate.  

# Discussion  

In this analysis, it was found that the levels of neural activity in the visual cortex are dependent on an interaction between the left and right contrast levels. However, despite some types of trials being notably easier than others, there were still quite a number of missed trials for said easier combinations. That said, the choice of analytical method for this project was a relatively crude and brute force method. The mean is an overly simplified measure of the neural activity and completely ignores any possible patterns that could correlate with certain decisions being made. Further analysis options for this data set would include a more in-depth attempt at distinguishing the neurons by type depending on the responses to certain stimuli. Perhaps actually looking at which types of combinations of stimuli lead to extremely high or low levels of activity would be useful. Or, looking at the patterns of firings based on the stimuli may show trends related to how the stimuli are processed in the visual cortex. This could also improve possible prediction models, using the given patterns of how certain neuronal groups fire to determine the outcome of a trial. However, in the interest of time and due to my own shortcomings, I have chosen not to perform such an analysis.     

The prediction model also has some downfalls. For starters, the prediction model is not generalizable to other sessions or trial types because it was trained for the first session only. As a consequence, it also suffers from a lack of training data. The number of observations that the model is trained on is only about 140, which is only about 10% of the observations that could have been used to train the data. Therefore, in order to build a prediction model that encompasses other sessions, it may be better to do a different kind of split for the data so the model is trained over all sessions. However, in the interest of having higher accuracy, I opted to work with the smaller data set. Additonal possible improvements to the prediction model could be to include the clusters as a factor and determining whether high levels of activity of a neuron of a certain cluster can help predict the outcome of the trial.  

Given my chosen analytical methods, however, some general observations can be made. For one, the patterns of the distributions in firing rates for sessions 1,2,and 3 varied more than Sessions 4 and 5. This could be due to the time between sessions 1 and 2 being longer than the time between sessions 4 and 5. To be more specific, session 1 (Cori) happens on December 14, 2016 and session 2 (also Cori) happens on December 17, 2016. Session 3 occurs on December 18, 2016. The sessions for Frossman (sessions 4 and 5) are also on different dates, but the gap between them is also only one day. However, because the mice were trained on the protocol beforehand and the experiments were designed to maintain high reward rates, it can be argued that the mice aren't expected to be learning or become unwilling to participate over the course of just a few days.Perhaps not performing the trials for a while led to diminished recall of the trained behavior and the mouse had to relearn the process as it went, returning to its original levels after the it had finished relearning the procedure. This hypothetical scenario comes from Fig.1 and Fig.2, in which, like the distributions for sessions 4 and 5, the distributions for sessions 1 and 3 are very similar to each other but session 2 is notably different. Of course, further and more in depth experiments regarding the effect on the extinction of behaviors on the neural response would need to be conducted to reach a definitive conclusion. Other kinds of additional research can look into the possible effects of the trial length as it was observed that the average firing rate of the neurons had a tendency to decrease as the trials went on.  


# Reference {-}

Glazewski S, Barth AL. Stimulus intensity determines experience-dependent modifications in neocortical neuron firing rates. The European journal of neuroscience. https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4331261/. Published December 26, 2014. Accessed March 8, 2023. 


Steinmetz, N.A., Zatka-Haas, P., Carandini, M. et al. Distributed coding of choice, action and engagement across the mouse brain. Nature 576, 266–273 (2019). https://doi.org/10.1038/s41586-019-1787-x



